Finally Tagless, Partially Evaluated

### 2007-06-12 tagless/tagless-final.tex
16 pages

The first interpreter evaluates an object term to its value in the
metalanguage.  The module below interprets each object\hyp language
operation as the corresponding metalanguage operation
\begin{code3}
module R = struct type ('c,'dv) repr = 'dv (* no wrappers *)
  let int  (x:int)  = x         let bool (b:bool) = b
  let lam  f        = f         let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self
  let add  e1 e2    = e1 + e2   let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then et () else ee () end
\end{code3}
%
As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding} with |R|%
%aplas-specific
: \texttt{module EXR = EX(R)}%
.
%% \begin{code}
%% module EXR = EX(R)
%% \end{code}
Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
% aplas: commented out
\begin{comment}
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
In OCaml, too, it is obvious to the compiler that
\end{comment}
%aplas: the following line is for aplas only
It is obvious to the compiler that
%
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
%aplas-specific
(The source code shows a total interpreter |L| that measures
the size of each object term).
%
\begin{proposition}
If an object term~$e$ encoded in the metalanguage has type~$t$,
then evaluating~$e$ in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~$t$.
\end{proposition}
Generalizing from~|R| to all interpreters, we have the following
broader and more useful, if also more obvious, proposition.
\begin{proposition}
  If an implementation of |Symantics| never gets stuck, then
  the type system of the object
  language is sound with respect to the dynamic semantics defined by
  that implementation.
\end{proposition}
These propositions follow immediately from the soundness of the
metalanguage's type system.
% aplas-commented
\begin{comment}
For variety, we show another interpreter, which measures the \emph{size}
of each object term, defined as the number of term
constructors. The following is slightly abbreviated code (see the
accompanying source code for the complete definition).
\begin{code}
module L = struct
  type ('c,'dv) repr = int
  let int (x:int)  = 1
  let lam f        = f 0 + 1
  let app e1 e2    = e1 + e2 + 1
  let fix f        = f 0 + 1
  let mul e1 e2    = e1 + e2 + 1
  let if_ eb et ee = eb + et () + ee () + 1
end
\end{code}
Now the expression
\begin{code}
let module E = EX(L) in E.test1 ()
\end{code}
evaluates to |3|. This interpreter is not only tagless but also
total. It ``evaluates'' even seemingly divergent terms like
\begin{code}
app (fix (fun self -> self)) (int 1)
\end{code}

@@

(d) \begin{code2}
.<fun x_1 -> (fun x_2 -> let rec self_3 = fun n_4 ->
   (fun x_5 -> if x_5 <= 0 then 1  else x_2 * (self_3 (x_5 + (-1))))
   n_4 in self_3) x_1 7>.
\end{code2}
\caption{The tagless staged interpreter \texttt{C}.}
\label{fig:interpreter-C}
\end{figure}
%
This is a straightforward staging of
|module R|.
This compiler produces
unoptimized code. For example, interpreting our |test1|%
%aplas-specific
, Fig.~\ref{fig:interpreter-C}(b),
% fig:interpreter-C part (b) was here
gives the code value |.<(fun x_6 -> x_6) true>.|
of inferred type |('c, bool) C.repr|.  Interpreting |testpowfix7|
with
%aplas-specific
Fig.~\ref{fig:interpreter-C}(c),
% fig:interpreter-C part (c) was here
gives a code value with many apparent $\beta$-redexes
%aplas-specific
Fig.~\ref{fig:interpreter-C}(d).
% fig:interpreter-C part (d) was here
This compiler does not incur
any interpretive overhead: the
code produced for $\fun{x}x$ is simply |fun x_6 -> x_6| and does not
  call the interpreter, unlike the recursive calls to |eval0| and
  |eval| in the |L e| lines in \S\ref{tagproblem}.
The resulting code obviously contains no tags and no pattern matching.
%aplas: the line is for aplas. The original paragraph commented out
The accompanying code shows the corresponding Haskell implemention. 
\begin{comment}
We have also implemented this compiler in Haskell. 
Since Haskell
has no (convenient, typed) staging facility, we had to emulate
it. To be precise, we defined a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|%
\footnote{\texttt{ByteCode} can be mapped to AST of Template Haskell
  (TH). The output of our compiler will then be assuredly type-correct TH.}
Whereas our representation of object terms uses HOAS,
our bytecode uses integer-named
variables to be realistic. 
We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|.
The implementation\footnote{The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \texttt{ByteCode} \emph{data type}.} is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code above. The
accompanying code gives the full details.
\end{comment}

@@

Roughly, a partial evaluator interprets each object term to either
a static (present-stage) term (using the evaluator~|R|) or
a dynamic (future-stage) term (using the compiler~|C|).  To
distinguish between static and dynamic terms, one might define
the Haskell data type 
%aplas-specific
\texttt{data P0 t = S0 (R t) $\mid$ E0 (C t)}.
We then start to define |P0| as an instance of the |Symantics| class.
Integer and boolean literals are immediate, present-stage
values. Addition yields a static term (using~|R|) if and only if both operands
are static; otherwise we extract the dynamic terms from the operands and
add them using~|C|.
Whereas |mul| and |leq| are as easy to define as |add|, we encounter
a problem with |if_|.  If the first argument to |if_| is a dynamic term
(type |C Bool|) and the second and third arguments are a static term
(type |R a|) and a dynamic term (type |C a|), then we need to convert
the static term to dynamic, but there is no polymorphic ``lift''
function, of type |a -> C a|, to send a value to the future stage
\cite{xi-guarded,WalidPOPL03}.
%aplas-else
\begin{comment}
\begin{code}
data P0 t = S0 (R t) | E0 (C t)
\end{code}
To extract a dynamic term from this type, we create the functions
\begin{code}
abstrI :: P0 Int -> C Int
abstrI (S0 r) = int (unR r)
abstrI (E0 c) = c

@@

Our |Symantics| class includes only separate lifting methods |bool| and
|int|, not a parametrically polymorphic lifting method, for good reason:
When compiling to a first-order target language such as machine code,
booleans, integers, and functions may well be represented differently.
Thus polymorphic lift cannot be compiled without intensional type
analysis.  To avoid the need for polymorphic lift, we turn to
\citearound{'s technique}\citet[see also \citealp{sumii-hybrid}]{asai-binding-time}:
build a dynamic term
alongside every static term.

@@

We switch to the Haskell data type
%aplas-specific
|data P1 t = P1 (Maybe (R t)) (C t)|
%aplas-else
\begin{comment}
\begin{code}
data P1 t = P1 (Maybe (R t)) (C t)
\end{code}
\end{comment}
so that a partially evaluated term |P1 t| always contains a dynamic
component and sometimes contains a static component.  The two
alternative constructors of a |Maybe| value, |Just| and |Nothing|,
tag each partially evaluated term with a phase: either present or
future.  This tag is not an object type tag: all pattern matching below
is exhaustive.  Because the future-stage component is always present, we
can now define the polymorphic function
%aplas-specific
|abstr1 (P1 _ dyn) = dyn| of the type
|abstr1 :: P1 t -> C t|
%aplas-else
\begin{comment}
\begin{code}
abstr1 :: P1 t -> C t
abstr1 (P1 _ dyn) = dyn
\end{code}
\end{comment}
to extract it without requiring polymorphic lift into~|C|.  We then try
to define the interpreter |P1|---and get as far as the first-order
constructs of our object language, including |if_|.
\begin{code3}
instance Symantics P1 where
  int  x = P1 (Just (int x)) (int x)
  add (P1 (Just n1) _) (P1 (Just n2) _) = int (unR (add n1 n2))
  add e1 e2 = P1 Nothing (add (abstr1 e1) (abstr1 e2))
  if_ (P1 (Just s) _) et ef = if unR s then et else ef
  if_ eb et ef = P1 Nothing (if_ (abstr1 eb) (abstr1 et) (abstr1 ef))
\end{code3}
When we come to functions, however, we stumble.  According to our
definition of~|P1|, a partially evaluated object function, such as the
identity $\fun{x}x$ embedded in Haskell as |lam (\x -> x)|\texttt{ ::
}|P1 (a -> a)|, consists of a dynamic part (type |C (a -> a)|) and
maybe a static part (type |R (a -> a)|).  The dynamic part is useful
when this function is passed to another function that is only
dynamically known, as in $\fun{k}k(\fun{x}x)$.  The static part is
useful when this function is applied to a static argument, as in
$(\fun{x}x)\True$.  Neither part, however, lets us \emph{partially}
evaluate the function, that is, compute as much as possible statically
when it is applied to a mix of static and dynamic inputs.  For example,
the partial evaluator should turn $\fun{n}(\fun{x}x)n$ into $\fun{n}n$
by substituting $n$ for~$x$ in the body of $\fun{x}x$ even though $n$ is
not statically known.  Even the same static function, applied to
different static arguments, can give both static and dynamic results: we
want to simplify $(\fun{y}x\times y)0$ to~$0$ but $(\fun{y}x\times y)1$
to~$x$.

To enable these simplifications, we delay binding-time analysis (BTA)
for a static function until it is applied, that is, until |lam f|
appears as the argument of |app|.  To do so, we have to incorporate |f|
as it is into the |P1| data structure: applying the type constructor
|P1| to a function type |a -> b| should yield one of
\begin{code3}
data P1 (a -> b) = S1 (P1 a -> P1 b) | E1 (C (a -> b))
data P1 (a -> b) = P1 (Maybe (P1 a -> P1 b)) (C (a -> b))
\end{code3}
That is, we need a nonparametric data type, something akin to
type-indexed functions and type-indexed types, which
\citet{oliveira-typecase} dub the \emph{typecase} design pattern.
%aplas-simplification
Thus typed partial evaluation, like typed CPS transformation,
inductively defines a map from source types to target types that
performs case distinction on the source type.
\begin{comment}

@@

Two common ways to provide typecase in Haskell are
GADTs and type-class functional dependencies
\cite{oliveira-typecase}.  These
methods are equivalent, and here we show the GADT way; |incope1.hs|
in the accompanying source code shows the latter.
%aplas: the original text is in comments below. Here's a greatly
% abbreaviated text
We introduce a GADT with four data constructors.
\begin{code3}
data P t where
  VI :: Int  -> P Int;                VB :: Bool -> P Bool
  VF :: (P a -> P b) -> P (a -> b);   E  :: C t -> P t
\end{code3}
The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
in~\S\ref{S:PE-lift}), and |E| builds dynamic terms (like |E0|).  However,
the type |P t| is no longer parametric in~|t|: the constructor |VF| takes an
operand of type |P a -> P b| rather than |a -> b|. As before, we define a
function |abstr| to extract a future-stage computation from a value of 
type |P t|. The implementations of |int|, |bool|, and |add| are like
in~\S\ref{S:PE-problem}.
The interpretation of |lam f = VF f| just wraps the HOAS function |f|. 
We can always compile |f| to a code value,
but we delay it to apply |f| to concrete arguments. The interpretation of
\begin{code3}
app (VF f) ea = f ea;  app (E f)  ea = E (app f (abstr ea))
\end{code3}
checks to see if |ef| is such a delayed
HOAS function |VF f|. If it is, we apply |f| to the
concrete argument |ea|, giving us a chance to perform static
computations (see example |testpowfix7| in~\S\ref{S:PE-solution}). If |ef| is a
dynamic value |E f|, we residualize.
This solution using GADTs works but is not quite satisfactory. First, it
cannot be ported to MetaOCaml as GADTs are unavailable there.  Second,
the problem of nonexhaustive pattern\hyp matching reappears in
|app|. Although Coq or logical
frameworks with canonical forms could prove pattern-matching
exhaustive, GHC cannot and issues warnings. Recall, our goal is to
make this exhaustiveness \emph{syntactically} apparent.

\begin{comment}
We introduce a GADT with four data constructors.
\begin{code3}
data P t where
  VI :: Int  -> P Int
  VB :: Bool -> P Bool
  VF :: (P a -> P b) -> P (a -> b)
  E  :: C t -> P t
\end{code3}
The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
in~\S\ref{S:PE-lift}), and |E| builds dynamic terms (like |E0|).  However,
the type |P t| is no longer parametric in~|t|: the constructor |VF| takes an
operand of type |P a -> P b| rather than |a -> b|. As before, we define a
function to extract a future-stage computation from a value of type |P t|.
\begin{code}
abstr :: P t -> C t
abstr (VI i) = int i
abstr (VB b) = bool b
abstr (VF f) = lam (abstr . f . E)
abstr (E x)  = x
\end{code}
The cases of this function |abstr| are type-indexed.  In particular, the |VF f|
case uses the method |lam| of the |C| interpreter to compile~|f|.

@@

\subsection{The ``final'' solution}
\label{S:PE-solution}
Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
would ideally like is to write
%aplas: inlining
\texttt{data P t = P (Maybe (repr\_pe t)) (C t)}
\begin{comment}
\begin{code}
data P t = P (Maybe (repr_pe t)) (C t)
\end{code}
\end{comment}
where |repr_pe| is the type function defined
% inductively because P below depends on repr_pe
by\vspace*{-0.8em}
\begin{code3}
repr_pe Int      = Int          repr_pe Bool     = Bool
repr_pe (a -> b) = P a -> P b
\end{code3}
\vspace*{-0.8em}
Although we can use type classes to define this type function
in Haskell, that is not portable to MetaOCaml. However,
these three typecase alternatives are already present in existing
methods of |Symantics|.
A simple and portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. Switching to MetaOCaml to demonstrate this
portability, we recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
constructor took two arguments |'c| and~|'dv|. We add an argument
|'sv| for the result of applying |repr_pe| to~|'dv|.
Figure~\ref{fig:ocaml} shows the new |Symantics| signature.
% aplas: abbreviated figure
\begin{figure*}
\begin{floatrule}
\begin{code}
module type Symantics = sig type ('c,'sv,'dv) repr
  val int : int  -> ('c,int,int) repr
  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x) -> ('c,'x,'da->'db) repr
  val app : ('c,'x,'da->'db) repr -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
  val fix : ('x->'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr, 'da->'db) repr as 'x)
  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val if_ : ('c,bool,bool) repr -> (unit->'x) -> (unit->'x) -> (('c,'sa,'da) repr as 'x)
end
\end{code}
\end{floatrule}
\caption{A (Meta)OCaml embedding of our object language that supports
  partial evaluation (\texttt{bool}, \texttt{mul},
\texttt{leq} are elided)}
\label{fig:ocaml}
\end{figure*}

@@

To illustrate how to add optimizations, we improve |add| (and |mul|,
elided) to simplify the generated code using the monoid (and ring)
structure of~|int|: not only is addition performed statically
(using~|R|) when both operands are statically known, but it is
eliminated when one operand is statically~$0$; similarly for
multiplication by~$0$ or~$1$.  Such algebraic simplifications are easy
to abstract over the specific domain (such as monoid or ring) where they
apply.  These simplifications and abstractions help a lot in a large
language with more base types and primitive operations.

@@

%aplas: abbreviated paragraph. The original is in the comments below
Given this implementation of~|P|, our running example
\texttt{let module E = EX(P) in E.test1 ()} evaluates to
\texttt{\{P.st = Some true; P.dy = .<true>.\}}
of type |('a, bool, bool) P.repr|.  Unlike with~|C| in~\S\ref{compiler},
a $\beta$-reduction has been statically performed to yield |true|.  More
interestingly, whereas |testpowfix7| compiles to a code value with many
$\beta$-redexes in~\S\ref{compiler}, the partial evaluation
\texttt{let module E = EX(P) in E.testpowfix7}
gives the desired result
\begin{code3}
{P.st = Some <fun>;
 P.dy = .<fun x_1 -> x_1 * (x_1 * (x_1 * (x_1 * (x_1 * (x_1 * x_1)))))>.}
\end{code3}

@@

We solve the problem of embedding a typed object language in a typed
metalanguage without using GADTs, dependent types, or a universal type.
Our family of interpreters include an evaluator, a compiler, a partial
evaluator, and CPS transformers.  It is patent that they never get stuck,
because we represent object types as metalanguage types.  This work
makes it safer and more efficient to embed domain\hyp specific languages
in practical metalanguages such as Haskell and ML\@.

>>>>

The first interpreter evaluates an object term to its value in the
metalanguage.  The module below interprets each object\hyp language
operation as the corresponding metalanguage operation
\vspace*{-0.8em}\begin{code3}
module R = struct type ('c,'dv) repr = 'dv (* no wrappers *)
  let int  (x:int)  = x         let bool (b:bool) = b
  let lam  f        = f         let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self
  let add  e1 e2    = e1 + e2   let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then et () else ee () end
\end{code3}
\vspace*{-0.8em}
%
As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding} with |R|%
%aplas-specific
: \texttt{module EXR = EX(R)}%
.
%% \begin{code}
%% module EXR = EX(R)
%% \end{code}
Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
% aplas: commented out
\begin{comment}
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
In OCaml, too, it is obvious to the compiler that
\end{comment}
%aplas: the following line is for aplas only
It is obvious to the compiler that
%
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
%aplas-specific
(The source code shows a total interpreter |L| that measures
the size of each object term).
%aplas-specific (cont)
We can also generalize from~|R| to all interpreters; the proof
of these propositions follow immediately from the soundness of the
metalanguage's type system.
%end aplas-specific
\begin{proposition}
If an object term~$e$ encoded in the metalanguage has type~$t$,
then evaluating~$e$ in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~$t$.
\end{proposition}
% Generalizing from~|R| to all interpreters, we have the following
% broader and more useful, if also more obvious, proposition.
\begin{proposition}
  If an implementation of |Symantics| never gets stuck, then
  the type system of the object
  language is sound with respect to the dynamic semantics defined by
  that implementation.
\end{proposition}
% These propositions follow immediately from the soundness of the
% metalanguage's type system.
% aplas-commented
\begin{comment}
For variety, we show another interpreter, which measures the \emph{size}
of each object term, defined as the number of term
constructors. The following is slightly abbreviated code (see the
accompanying source code for the complete definition).
\begin{code}
module L = struct
  type ('c,'dv) repr = int
  let int (x:int)  = 1
  let lam f        = f 0 + 1
  let app e1 e2    = e1 + e2 + 1
  let fix f        = f 0 + 1
  let mul e1 e2    = e1 + e2 + 1
  let if_ eb et ee = eb + et () + ee () + 1
end
\end{code}
Now the expression
\begin{code}
let module E = EX(L) in E.test1 ()
\end{code}
evaluates to |3|. This interpreter is not only tagless but also
total. It ``evaluates'' even seemingly divergent terms like
\begin{code}
app (fix (fun self -> self)) (int 1)
\end{code}

@@

(d) \begin{code2}
.<fun x_1 -> (fun x_2 -> let rec self_3 = fun n_4 ->
   (fun x_5 -> if x_5 <= 0 then 1  else x_2 * (self_3 (x_5 + (-1))))
   n_4 in self_3) x_1 7>.
\end{code2}
\caption{The tagless staged interpreter \texttt{C}.}
\label{fig:interpreter-C}
\end{figure}
%
This is a straightforward staging of
|module R|.
This compiler produces
unoptimized code. For example, interpreting our |test1|%
%aplas-specific
, Fig.~\ref{fig:interpreter-C}(b),
% fig:interpreter-C part (b) was here
gives the code value |.<(fun x_6 -> x_6) true>.|
of inferred type |('c, bool) C.repr|.  Interpreting |testpowfix7|
with
%aplas-specific
Fig.~\ref{fig:interpreter-C}(c),
% fig:interpreter-C part (c) was here
gives a code value with many apparent $\beta$-redexes
%aplas-specific
Fig.~\ref{fig:interpreter-C}(d).
% fig:interpreter-C part (d) was here
This compiler does not incur
any interpretive overhead: the
code produced for $\fun{x}x$ is simply |fun x_6 -> x_6| 
% aplas: and does not
% call the interpreter, unlike the recursive calls to |eval0| and
% |eval| in the |L e| lines in \S\ref{tagproblem}.
The resulting code obviously contains no tags and no pattern matching.
%aplas: the line is for aplas. The original paragraph commented out
The accompanying code shows the Haskell implementation. 
\begin{comment}
We have also implemented this compiler in Haskell. 
Since Haskell
has no (convenient, typed) staging facility, we had to emulate
it. To be precise, we defined a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|%
\footnote{\texttt{ByteCode} can be mapped to AST of Template Haskell
  (TH). The output of our compiler will then be assuredly type-correct TH.}
Whereas our representation of object terms uses HOAS,
our bytecode uses integer-named
variables to be realistic. 
We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|.
The implementation\footnote{The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \texttt{ByteCode} \emph{data type}.} is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code above. The
accompanying code gives the full details.
\end{comment}

@@

Roughly, a partial evaluator interprets each object term to either
a static (present-stage) term (using~|R|) or
a dynamic (future-stage) term (using~|C|).  To
distinguish between static and dynamic terms, one might define
%aplas-specific
\texttt{data P0 t = S0 (R t) $\mid$ E0 (C t)}.
We then start to define |P0| as an instance of the |Symantics| class.
Integer and boolean literals are immediate, present-stage
values. Addition yields a static term (using~|R|) if and only if both operands
are static; otherwise we extract the dynamic terms from the operands and
add them using~|C|.
Whereas |mul| and |leq| are as easy to define as |add|, we encounter
a problem with |if_|.  If the first argument to |if_| is a dynamic term
(type |C Bool|) and the second and third arguments are a static term
(type |R a|) and a dynamic term (type |C a|), then we need to convert
the static term to dynamic, but there is no polymorphic ``lift''
function, of type |a -> C a|, to send a value to the future stage
\cite{xi-guarded,WalidPOPL03}.
%aplas-else
\begin{comment}
\begin{code}
data P0 t = S0 (R t) | E0 (C t)
\end{code}
To extract a dynamic term from this type, we create the functions
\begin{code}
abstrI :: P0 Int -> C Int
abstrI (S0 r) = int (unR r)
abstrI (E0 c) = c

@@

Our |Symantics| class only includes separate lifting methods |bool| and
|int|, not a parametrically polymorphic lifting method:
When compiling to a first-order target language such as machine code,
booleans, integers, and functions may well be represented differently.
Thus polymorphic lift cannot be compiled without intensional type
analysis.  To avoid this, we turn to
\citearound{'s technique}\citet[see also \citealp{sumii-hybrid}]{asai-binding-time}:
build a dynamic term
alongside every static term.

@@

We switch to the Haskell data type
%aplas-specific
|data P1 t = P1 (Maybe (R t)) (C t)|
%aplas-else
\begin{comment}
\begin{code}
data P1 t = P1 (Maybe (R t)) (C t)
\end{code}
\end{comment}
so that a partially evaluated term |P1 t| always contains a dynamic
component and sometimes contains a static component.  The two
alternative constructors of a |Maybe| value, |Just| and |Nothing|,
tag each term with a phase: present or
future.  This tag is not an object type tag: all pattern matching below
is exhaustive.  Because the future-stage component is always present, we
can now define the polymorphic function
%aplas-specific
|abstr1 (P1 _ dyn) = dyn| of type
|abstr1 :: P1 t -> C t|
%aplas-else
\begin{comment}
\begin{code}
abstr1 :: P1 t -> C t
abstr1 (P1 _ dyn) = dyn
\end{code}
\end{comment}
to extract it without requiring polymorphic lift into~|C|.  We then try
to define the interpreter |P1|---and get as far as the first-order
constructs of our object language, including |if_|.
\vspace*{-0.7em}
\begin{code3}
instance Symantics P1 where
  int  x = P1 (Just (int x)) (int x)
  add (P1 (Just n1) _) (P1 (Just n2) _) = int (unR (add n1 n2))
  add e1 e2 = P1 Nothing (add (abstr1 e1) (abstr1 e2))
  if_ (P1 (Just s) _) et ef = if unR s then et else ef
  if_ eb et ef = P1 Nothing (if_ (abstr1 eb) (abstr1 et) (abstr1 ef))
\end{code3}
\vspace*{-0.7em}
When we come to functions, however, we stumble.  According to our
definition of~|P1|, a partially evaluated object function, such as the
identity $\fun{x}x$ embedded in Haskell as |lam (\x -> x)|\texttt{ ::
}|P1 (a -> a)|, consists of a dynamic part (type |C (a -> a)|) and
maybe a static part (type |R (a -> a)|).  The dynamic part is useful
when this function is passed to another function that is only
dynamically known, as in $\fun{k}k(\fun{x}x)$.  The static part is
useful when this function is applied to a static argument, as in
$(\fun{x}x)\True$.  Neither part, however, lets us \emph{partially}
evaluate the function, that is, compute as much as possible statically
when it is applied to a mix of static and dynamic inputs.  For example,
the partial evaluator should turn $\fun{n}(\fun{x}x)n$ into $\fun{n}n$
by substituting $n$ for~$x$ in the body of $\fun{x}x$ even though $n$ is
not statically known.  Even the same static function, applied to
different static arguments, can give both static and dynamic results: we
want to simplify $(\fun{y}x\times y)0$ to~$0$ but $(\fun{y}x\times y)1$
to~$x$.

To enable these simplifications, we delay binding-time analysis (BTA)
for a static function until it is applied, that is, until |lam f|
appears as the argument of |app|.  To do so, we have to incorporate |f|
as it is into the |P1| data structure: applying the type constructor
|P1| to a function type |a -> b| should yield one of
\vspace*{-0.7em}
\begin{code3}
data P1 (a -> b) = S1 (P1 a -> P1 b) | E1 (C (a -> b))
data P1 (a -> b) = P1 (Maybe (P1 a -> P1 b)) (C (a -> b))
\end{code3}
\vspace*{-0.7em}
That is, we need a nonparametric data type, something akin to
type-indexed functions and type-indexed types, which
\citet{oliveira-typecase} dub the \emph{typecase} design pattern.
%aplas-simplification
Thus typed partial evaluation, like typed CPS transformation,
inductively defines a map from source types to target types that
performs case distinction on the source type.
\begin{comment}

@@

Two common ways to provide typecase in Haskell are
GADTs and type-class functional dependencies
\cite{oliveira-typecase}.  These
methods are equivalent, and here we use GADTs; |incope1.hs|
in the accompanying source code shows the latter.
%aplas: the original text is in comments below. Here's a greatly
% abbreaviated text
\vspace*{-0.7em}
\begin{code3}
data P t where
  VI :: Int  -> P Int;                VB :: Bool -> P Bool
  VF :: (P a -> P b) -> P (a -> b);   E  :: C t -> P t
\end{code3}
\vspace*{-0.7em}
The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
in~\S\ref{S:PE-lift}), and |E| builds dynamic terms (like |E0|).  However,
the type |P t| is no longer parametric in~|t|: the constructor |VF| takes an
operand of type |P a -> P b| rather than |a -> b|. As before, we define a
function |abstr| to extract a future-stage computation from a value of 
type |P t|. The implementations of |int|, |bool|, and |add| are like
in~\S\ref{S:PE-problem}.
The interpretation of |lam f = VF f| just wraps the HOAS function |f|. 
We can always compile |f| to a code value,
but we delay it to apply |f| to concrete arguments. The interpretation of
\vspace*{-0.7em}
\begin{code3}
app (VF f) ea = f ea;  app (E f)  ea = E (app f (abstr ea))
\end{code3}
\vspace*{-0.7em}
checks to see if |ef| is such a delayed
HOAS function |VF f|. If it is, we apply |f| to the
concrete argument |ea|, giving us a chance to perform static
computations (see example |testpowfix7| in~\S\ref{S:PE-solution}). If |ef| is a
dynamic value |E f|, we residualize.
This solution using GADTs works but is not quite satisfactory. First, it
cannot be ported to MetaOCaml.  Second,
the problem of nonexhaustive pattern\hyp matching reappears in
|app|. Although Coq or logical
frameworks with canonical forms could prove pattern-matching
exhaustive, GHC cannot and issues warnings. Recall, our goal is to
make this exhaustiveness \emph{syntactically} apparent.

\begin{comment}
We introduce a GADT with four data constructors.
\vspace*{-0.7em}
data P t where
  VI :: Int  -> P Int
  VB :: Bool -> P Bool
  VF :: (P a -> P b) -> P (a -> b)
  E  :: C t -> P t
\end{code3}
The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
in~\S\ref{S:PE-lift}), and |E| builds dynamic terms (like |E0|).  However,
the type |P t| is no longer parametric in~|t|: the constructor |VF| takes an
operand of type |P a -> P b| rather than |a -> b|. As before, we define a
function to extract a future-stage computation from a value of type |P t|.
\begin{code}
abstr :: P t -> C t
abstr (VI i) = int i
abstr (VB b) = bool b
abstr (VF f) = lam (abstr . f . E)
abstr (E x)  = x
\end{code}
The cases of this function |abstr| are type-indexed.  In particular, the |VF f|
case uses the method |lam| of the |C| interpreter to compile~|f|.

@@

\subsection{The ``final'' solution}
\label{S:PE-solution}
Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
would ideally like is to write
%aplas: inlining
\texttt{data P t = P (Maybe (repr\_pe t)) (C t)}
\begin{comment}
\begin{code}
data P t = P (Maybe (repr_pe t)) (C t)
\end{code}
\end{comment}
where |repr_pe| is the type function defined
% inductively because P below depends on repr_pe
by\vspace*{-0.8em}
\begin{code3}
repr_pe Int      = Int          repr_pe Bool     = Bool
repr_pe (a -> b) = P a -> P b
\end{code3}
\vspace*{-0.8em}
Although we can use type classes to define this type function
in Haskell, that is not portable to MetaOCaml. However,
these three typecase alternatives are already present in existing
methods of |Symantics|.
A simple and portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. Switching to MetaOCaml to demonstrate this,
we recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
constructor took two arguments |'c| and~|'dv|. We add an argument
|'sv| for the result of applying |repr_pe| to~|'dv|.
Figure~\ref{fig:ocaml} shows the new signature.
% aplas: abbreviated figure
\begin{figure*}
\begin{floatrule}
\begin{code}
module type Symantics = sig type ('c,'sv,'dv) repr
  val int : int  -> ('c,int,int) repr
  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x) -> ('c,'x,'da->'db) repr
  val app : ('c,'x,'da->'db) repr -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
  val fix : ('x->'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr, 'da->'db) repr as 'x)
  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val if_ : ('c,bool,bool) repr -> (unit->'x) -> (unit->'x) -> (('c,'sa,'da) repr as 'x)
end
\end{code}
\end{floatrule}
\caption{A (Meta)OCaml embedding of our object language that supports
  partial evaluation (\texttt{bool}, \texttt{mul},
\texttt{leq} are elided)}
\label{fig:ocaml}
\end{figure*}

@@

To illustrate how to add optimizations, we improve |add| (and |mul|,
elided) to simplify the generated code using the monoid (and ring)
structure of~|int|: not only is addition performed statically
(using~|R|) when both operands are statically known, but it is
eliminated when one operand is statically~$0$; similarly for
multiplication by~$0$ or~$1$.  
\begin{comment}
Such algebraic simplifications are easy
to abstract over the specific domain (such as monoid or ring) where they
apply.  These simplifications and abstractions help a lot in a large
language with more base types and primitive operations.
\end{comment}
%aplas-specific
In large languages, this can be quite effective.

@@

%aplas: abbreviated paragraph. The original is in the comments below
Given this implementation of~|P|, our running example
\texttt{let module E = EX(P) in E.test1 ()} evaluates to
\texttt{\{P.st = Some true; P.dy = .<true>.\}}
of type |('a, bool, bool) P.repr|.  Unlike with~|C| in~\S\ref{compiler},
a $\beta$-reduction has been statically performed to yield |true|.  More
interestingly, whereas |testpowfix7| compiles to a code value with many
$\beta$-redexes in~\S\ref{compiler}, the partial evaluation
\texttt{let module E = EX(P) in E.testpowfix7}
gives the desired result
\vspace*{-0.7em}
\begin{code3}
{P.st = Some <fun>;
 P.dy = .<fun x_1 -> x_1 * (x_1 * (x_1 * (x_1 * (x_1 * (x_1 * x_1)))))>.}
\end{code3}
\vspace*{-0.7em}

@@

We solve the problem of embedding a typed object language in a typed
metalanguage without using GADTs, dependent types, or a universal type.
Our family of interpreters include an evaluator, a compiler, a partial
evaluator, and CPS transformers.  It is patent that they never get stuck,
because we represent object types as metalanguage types.  This work
makes it safer and more efficient to embed %domain\hyp specific languages
DSLs
in practical metalanguages such as Haskell and ML\@.
